Variance reduction for MCMC methods via martingale representations

### 2020-03-03 aap/Revision/vr-mcmc-revision-new.tex
update

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\[
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
\]
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step size $\gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

>>>>

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\) and \(f(x) = \sum_{i=1}^d x^2_i\) we have used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent "training" trajectories of length $n = 10^4$.  Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

\subsection{Binary Logistic Regression}

Second experiment considers the problem of logistic regression. Let $\mathsf{Y} = \left(\mathsf{Y}_1,\ldots,\mathsf{Y}_n\right) \in \{0,1\}^m$ be binary response variables, $\mathsf{X} \in \rset^{m \times d}$ be a feature matrix and $\theta \in \rset^d$ - vector of regression parameters. We define log-likelihood of $i$-th observation as
\[
\ell(\mathsf{Y}_i | \theta, \mathsf{X}_i) = \mathsf{Y}_i \mathsf{X}_i^T\theta - \ln{\left(1 + \rme^{\mathsf{X}_i^T\theta}\right)}
\]
In order to estimate $\theta$ according to given data, the Bayesian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta| \mathsf{Y}, \mathsf{X})$ using Bayes' rule.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{../pictures/logreg_1st_order.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
 \label{fig:2}}
\end{center}
\end{figure}

In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
\begin{eqnarray*}
\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
\end{eqnarray*}
Thus we obtain
\begin{eqnarray*}
U(\theta) = \mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{-\theta^T \mathsf{X}_i})} + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories. We put $n_{trunc} = 50$ and $K = 1$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$. In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
